---
title: "Visualizing models"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4)
library(coursekata)
```

A model is a claim about what value to predict. `gf_model()` draws that claim on
top of the data it was fit to, so the claim and the evidence are in the same
picture.

## The empty model

The empty model predicts the same value — the mean — for every observation. On a
histogram it is a single vertical line.

```{r}
empty_model <- lm(body_mass_kg ~ NULL, data = penguins)
gf_histogram(~body_mass_kg, data = penguins, binwidth = 0.25) %>%
  gf_model(empty_model)
```

## Group models

With a categorical explanatory variable the model predicts one value per group,
and `gf_model()` draws a mark at each group mean.

```{r}
species_model <- lm(body_mass_kg ~ species, data = penguins)
gf_jitter(body_mass_kg ~ species, data = penguins, width = .1) %>%
  gf_model(species_model)
```

The same model layers onto faceted histograms, which makes the comparison
between groups a comparison between panels.

```{r fig.height = 6}
gf_histogram(~body_mass_kg, data = penguins, binwidth = 0.25) %>%
  gf_facet_grid(species ~ .) %>%
  gf_model(species_model)
```

## Regression models

With a quantitative explanatory variable the prediction is a line.

```{r}
flipper_model <- lm(body_mass_kg ~ flipper_length_m, data = penguins)
gf_point(body_mass_kg ~ flipper_length_m, data = penguins) %>%
  gf_model(flipper_model)
```

Layering two models in different colors puts the comparison on one plot: the
empty model ignores flipper length, the regression model uses it.

```{r}
gf_point(body_mass_kg ~ flipper_length_m, data = penguins) %>%
  gf_model(empty_model, color = "dodgerblue") %>%
  gf_model(flipper_model, color = "firebrick")
```

## Residuals

A residual is the distance from what the model predicted to what actually
happened. `gf_resid()` draws them as segments. With hundreds of points this is
unreadable, so take a small sample first.

```{r}
set.seed(1)
penguins_20 <- sample(penguins, 20)
sample_model <- lm(body_mass_kg ~ flipper_length_m, data = penguins_20)

gf_point(body_mass_kg ~ flipper_length_m, data = penguins_20) %>%
  gf_model(sample_model) %>%
  gf_resid(sample_model, color = "firebrick")
```

`gf_square_resid()` draws each residual as a square instead of a segment, which
makes *squared* error visible as area — the quantity least squares is actually
minimizing.

```{r}
gf_point(body_mass_kg ~ flipper_length_m, data = penguins_20) %>%
  gf_model(sample_model) %>%
  gf_square_resid(sample_model, color = "firebrick")
```

Comparing the empty model's squares to the regression model's is the clearest
picture of what the predictor bought you.

```{r}
empty_20 <- lm(body_mass_kg ~ NULL, data = penguins_20)
gf_point(body_mass_kg ~ flipper_length_m, data = penguins_20) %>%
  gf_model(empty_20) %>%
  gf_square_resid(empty_20, color = "dodgerblue")
```

## The standard deviation as a typical residual

`gf_sd_ruler()` draws one standard deviation, anchored at the mean. Under the
empty model the residuals *are* deviations from the mean, so the ruler is a
picture of a typical one.

```{r}
gf_point(Thumb ~ Height, data = Fingers, alpha = .4) %>%
  gf_model(lm(Thumb ~ NULL, data = Fingers)) %>%
  gf_sd_ruler(where = "mean")
```

On a histogram the outcome is on the x-axis, so the ruler turns horizontal and
runs along the baseline from the mean to one SD above it.

```{r}
gf_histogram(~Thumb, data = Fingers, binwidth = 5) %>%
  gf_model(lm(Thumb ~ NULL, data = Fingers)) %>%
  gf_sd_ruler(color = "red", size = 2)
```

Two groups with the same mean and different spread get rulers of different
lengths, which is the comparison the ruler exists to make. Fixing the axis with
`gf_lims()` keeps the two plots on the same scale.

```{r}
set.seed(154)
no_feedback <- data.frame(median_time = round(rnorm(100, 13, 6), 1))
set.seed(141)
feedback <- data.frame(median_time = round(rnorm(100, 13, 3), 1))

gf_histogram(~median_time, data = no_feedback, binwidth = 1) %>%
  gf_lims(x = c(0, 31)) %>%
  gf_model(lm(median_time ~ NULL, data = no_feedback)) %>%
  gf_sd_ruler(color = "red", size = 2)

gf_histogram(~median_time, data = feedback, binwidth = 1) %>%
  gf_lims(x = c(0, 31)) %>%
  gf_model(lm(median_time ~ NULL, data = feedback)) %>%
  gf_sd_ruler(color = "red", size = 2)
```
