---
title: "Visualizing models"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4)
library(coursekata)
```

A model is a claim about what value to predict. `gf_model()` draws that claim on
top of the data it was fit to, so the claim and the evidence are in the same
picture.

## The empty model

The empty model predicts the same value, the mean, for every observation. On a
histogram it is a single vertical line.

```{r}
empty_model <- lm(body_mass_kg ~ NULL, data = penguins)
gf_histogram(~body_mass_kg, data = penguins, binwidth = 0.25) %>%
  gf_model(empty_model)
```

A plot already implies a model, so you can leave the argument out and let
`gf_model()` work out which one. A distribution on its own implies the empty model.

```{r}
gf_histogram(~body_mass_kg, data = penguins, binwidth = 0.25) %>%
  gf_model()
```

## Group models

With a categorical explanatory variable the model predicts one value per group,
and `gf_model()` draws a mark at each group mean.

```{r}
species_model <- lm(body_mass_kg ~ species, data = penguins)
gf_jitter(body_mass_kg ~ species, data = penguins, width = .1) %>%
  gf_model(species_model)
```

The same model layers onto faceted histograms, which makes the comparison
between groups a comparison between panels.

```{r fig.height = 6}
gf_histogram(~body_mass_kg, data = penguins, binwidth = 0.25) %>%
  gf_facet_grid(species ~ .) %>%
  gf_model(species_model)
```

## Regression models

With a quantitative explanatory variable the prediction is a line.

```{r}
flipper_model <- lm(body_mass_kg ~ flipper_length_m, data = penguins)
gf_point(body_mass_kg ~ flipper_length_m, data = penguins) %>%
  gf_model(flipper_model)
```

Layering two models in different colors puts the comparison on one plot: the
empty model ignores flipper length, the regression model uses it.

```{r}
gf_point(body_mass_kg ~ flipper_length_m, data = penguins) %>%
  gf_model(empty_model, color = "dodgerblue") %>%
  gf_model(flipper_model, color = "firebrick")
```

## The ggplot2 front door

`geom_model()` draws the same model layer on an ordinary ggplot2 plot. Supply a
fitted model or a two-sided formula.

```{r}
ggplot(penguins, aes(flipper_length_m, body_mass_kg)) +
  geom_point() +
  geom_model(model = flipper_model)
```

Leave `model` out to fit the model represented by the layer's mappings. As with
other ggplot2 stats, this is computed separately in each panel and group.

```{r}
ggplot(penguins, aes(species, body_mass_kg)) +
  geom_jitter(width = .1) +
  geom_model()
```

`stat_model()` exposes the same calculation when you want to choose the drawing
geom separately. `gf_model()` provides the same model behavior with ggformula
syntax.

## Residuals

A residual is the observed value minus the model's prediction. `gf_resid()`
draws a segment between those two values. With hundreds of points the segments
are hard to read, so take a small sample first.

```{r}
set.seed(1)
penguins_20 <- sample(penguins, 20)
sample_model <- lm(body_mass_kg ~ flipper_length_m, data = penguins_20)

gf_point(body_mass_kg ~ flipper_length_m, data = penguins_20) %>%
  gf_model(sample_model) %>%
  gf_resid(sample_model, color = "firebrick")
```

The ggplot2 layer uses the same implementation. Give it the fitted model;
the plot still supplies the observations and their mappings.

```{r}
ggplot(penguins_20, aes(flipper_length_m, body_mass_kg)) +
  geom_point() +
  geom_resid(model = sample_model, color = "firebrick")
```

For jittered points, use one seeded position for the points and residuals. The
points move; the fitted ends of the residuals stay on the model.

```{r}
jitter <- position_jitter(width = .1, seed = 42)
gentoo_model <- lm(body_mass_kg ~ gentoo, data = penguins_20)

ggplot(penguins_20, aes(gentoo, body_mass_kg)) +
  geom_point(position = jitter) +
  geom_resid(model = gentoo_model, position = jitter, color = "firebrick")
```

`gf_square_resid()` draws each residual as a square instead of a segment, which
makes *squared* error visible as area: the quantity least squares is actually
minimizing.

```{r}
gf_point(body_mass_kg ~ flipper_length_m, data = penguins_20) %>%
  gf_model(sample_model) %>%
  gf_square_resid(sample_model, color = "firebrick")
```

Comparing the empty model's squares to the regression model's is the clearest
picture of what the predictor bought you.

```{r}
empty_20 <- lm(body_mass_kg ~ NULL, data = penguins_20)
gf_point(body_mass_kg ~ flipper_length_m, data = penguins_20) %>%
  gf_model(empty_20) %>%
  gf_square_resid(empty_20, color = "dodgerblue")
```

## Where the error goes

The blue squares show the error left by the empty model. The red squares show
the error left after the model uses flipper length. `gf_reduce()` draws the
distance between the two predictions: for each penguin, from the grand mean to
the regression model's prediction.

```{r}
gf_point(body_mass_kg ~ flipper_length_m, data = penguins_20) %>%
  gf_model(empty_20) %>%
  gf_model(sample_model) %>%
  gf_reduce(sample_model, color = "forestgreen")
```

Across the whole sample, the blue square areas add up to the red areas plus the
green areas. This identity holds for the sums, not for each penguin's three
squares on its own.

```{r}
gf_point(body_mass_kg ~ flipper_length_m, data = penguins_20) %>%
  gf_model(sample_model) %>%
  gf_square_resid(empty_20, color = "dodgerblue") %>%
  gf_square_resid(sample_model, color = "firebrick") %>%
  gf_square_reduce(sample_model, color = "forestgreen")
```

The direct ggplot2 counterparts are `geom_square_resid()`, `geom_reduce()`, and
`geom_square_reduce()`. The `stat_*()` functions separate the statistical and
drawing choices: `stat_resid()` and `stat_reduce()` default to segments, while
`geom = "square_resid"` draws the squared quantities as areas.

```{r}
ggplot(penguins_20, aes(flipper_length_m, body_mass_kg)) +
  geom_point() +
  stat_resid(model = empty_20, geom = "square_resid", color = "dodgerblue") +
  stat_resid(model = sample_model, geom = "square_resid", color = "firebrick") +
  stat_reduce(model = sample_model, geom = "square_resid", color = "forestgreen")
```

The sum of the green areas divided by the sum of the blue areas is PRE, the same
number `supernova()` puts in a table. Reduction layers require an unweighted
model with an intercept because that is what makes the three sums of squares
add up.

```{r}
supernova(sample_model)
```

## The standard deviation as a typical residual

`gf_sd_ruler()` draws one standard deviation, anchored at the mean. Under the
empty model the residuals *are* deviations from the mean, so the ruler is a
picture of a typical one.

```{r}
gf_point(Thumb ~ Height, data = Fingers, alpha = .4) %>%
  gf_model(lm(Thumb ~ NULL, data = Fingers)) %>%
  gf_sd_ruler(where = "mean")
```

On a histogram the outcome is on the x-axis, so the ruler turns horizontal and
runs along the baseline from the mean to one SD above it.

```{r}
gf_histogram(~Thumb, data = Fingers, binwidth = 5) %>%
  gf_model(lm(Thumb ~ NULL, data = Fingers)) %>%
  gf_sd_ruler(color = "red", linewidth = 2)
```

Two groups with the same mean and different spread get rulers of different
lengths, which is the comparison the ruler exists to make. Zooming both to the
same x range keeps them comparable. `coord_cartesian()` does that by changing
what you see; `gf_lims()` would do it by discarding anything outside the range,
which is not what you want when the whole point is to compare distributions.

```{r}
set.seed(154)
no_feedback <- data.frame(median_time = round(rnorm(100, 13, 6), 1))
set.seed(141)
feedback <- data.frame(median_time = round(rnorm(100, 13, 3), 1))

gf_histogram(~median_time, data = no_feedback, binwidth = 1) %>%
  gf_refine(coord_cartesian(xlim = c(0, 31))) %>%
  gf_model(lm(median_time ~ NULL, data = no_feedback)) %>%
  gf_sd_ruler(color = "red", linewidth = 2)

gf_histogram(~median_time, data = feedback, binwidth = 1) %>%
  gf_refine(coord_cartesian(xlim = c(0, 31))) %>%
  gf_model(lm(median_time ~ NULL, data = feedback)) %>%
  gf_sd_ruler(color = "red", linewidth = 2)
```
