---
title: "Sampling distributions"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4)
library(coursekata)
```

A sampling distribution answers a question about *what else could have
happened*. If the empty model were true — if condition made no difference to
tips at all — how big a `b1` would we see just from the luck of which tables
happened to land in which group?

`do()` answers that by running the same estimate many times over data that has
been shuffled, so any relationship between the two variables is broken by
construction.

```{r}
set.seed(42)
sdob1 <- do(1000) * b1(shuffle(Tip) ~ Condition, data = TipExperiment)
head(sdob1)
```

Each row is one `b1` from one shuffle. Plotted, they are the range of estimates
the empty model produces.

```{r}
gf_histogram(~b1, data = sdob1, binwidth = 1)
```

## Marking off regions

The distribution part functions turn a region of that distribution into a fill
aesthetic. `middle()` marks the values we would expect to see often.

```{r}
gf_histogram(~b1, data = sdob1, fill = ~ middle(b1, .95), binwidth = 1)
```

`tails()` marks the same cutoffs with the opposite coloring — the 5% most
extreme values, the ones that would be surprising under the empty model.

```{r}
gf_histogram(~b1, data = sdob1, fill = ~ tails(b1, .95), binwidth = 1)
```

`outer()` marks the same region but takes the tail proportion directly, which
reads more naturally when you are thinking in terms of how much is *out* rather
than how much is *in*.

```{r}
gf_histogram(~b1, data = sdob1, fill = ~ outer(b1, .05), binwidth = 1)
```

For a directional hypothesis, all of the 5% goes in one tail.

```{r}
gf_histogram(~b1, data = sdob1, fill = ~ upper(b1, .05), binwidth = 1)
gf_histogram(~b1, data = sdob1, fill = ~ lower(b1, .05), binwidth = 1)
```

## Bootstrapping instead of shuffling

Swap `shuffle()` for `resample()` and the same pattern produces a bootstrap
confidence interval rather than a null distribution. Nothing is shuffled, so the
relationship stays intact; what varies is which observations are drawn.

```{r}
set.seed(42)
sdob1_boot <- do(1000) * b1(Tip ~ Condition, data = resample(TipExperiment))
gf_histogram(~b1, data = sdob1_boot, fill = ~ middle(b1, .95), bins = 100)
```

## Framing the distribution with its DGP

`gf_squareplot()` draws the same distribution as countable squares, and
`show_dgp()` frames it with the data generating process: the population
model on the top axis, the sample estimate on the bottom, and a marker at the
null hypothesis. Both overlays take a plot and give one back, so the same
frame goes just as readily on a `gf_histogram()` of the same shuffles.

With only ten shuffles the mean of the distribution can land well away from the
null.

```{r}
set.seed(42)
small <- do(10) * b1(shuffle(Tip) ~ Condition, data = TipExperiment)
gf_squareplot(~b1, data = small, binwidth = 2) %>%
  gf_lims(x = c(-30, 30)) %>%
  gf_refine(expand_limits(y = 10)) %>%
  show_mean() %>%
  show_dgp()
```

With a hundred it settles close to the null, which is what the empty model
predicts. `expand_limits(y = )` sets a floor under the count axis, so a
distribution with a low peak still gets a full-height panel; a hundred
shuffles stack higher than ten and the axis grows to fit them.

```{r}
set.seed(42)
larger <- do(100) * b1(shuffle(Tip) ~ Condition, data = TipExperiment)
gf_squareplot(~b1, data = larger, binwidth = 2) %>%
  gf_lims(x = c(-30, 30)) %>%
  gf_refine(expand_limits(y = 10)) %>%
  show_mean() %>%
  show_dgp()
```
